Mergeable Types

### 2017-02-26 paper/mergeable-types.tex

The definition asserts the existence of two functions for a mergeable
type $t$. First, the function $W$, which computes the \emph{edit
distance} between an object $A:t$ and its semantic successor $B:t$.
The sequence of primitive morphisms it returns ($P^*$) is called an
\emph{edit script}. Replaying the edit script on $A$ produces $B$
(i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
existence of a function $M$, which performs an \emph{operational
transformation} of the primitive morphisms in an edit script. An
operational transformation of a primitive morphism $P$ w.r.t a
morphism $Q$ is another primitive morphism $P'$ such that for any
object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
$A$. We will now consider examples of mergeable types that reify the
notions of edit scripts and operational transformations.
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.4]{Figures/string-eg}
\caption{Merging strings with edit scripts and operational
transformations}
\label{fig:string-eg}
\end{wrapfigure}
\paragraph{Strings} Strings are mergeable. Given a string ($s$) and
its semantic successor ($s'$), an edit script ($P^*$ or $Q^*$) is the
shortest sequence of \C{insert}, \C{delete}, and \C{substitute}
operations (a string's primitive morphisms), whose composition takes $s$
to $s'$. The length of the sequence is in fact the Levenshtein
distance~\cite{levenstein} between $s$ and $s'$, which can be computed
in polynomial time~\cite{wagner-fischer}.
A slight modification of Wagner-Fischer also lets us compute the edit
script. Fig.~\ref{fig:string-eg} illustrates.

Next, we apply operational transformation to compute $P^*$ given $Q^*$
($P^* <+ Q^*$), \emph{or} $Q^*$ given $P^*$ ($Q^* <+ P^*$); there is
no need to compute both unless verification is one of the aims.

\paragraph{}

>>>>

The definition asserts the existence of two functions for a mergeable
type $t$. First, the function $W$, which computes the \emph{edit
distance} between an object $A:t$ and its semantic successor $B:t$.
The sequence of primitive morphisms it returns ($P^*$) is called an
\emph{edit script}. Replaying the edit script on $A$ produces $B$
(i.e., $\denot{P^*}(A) = B$). Second, the definition asserts the
existence of a function $M$, which performs an \emph{operational
transformation} of the primitive morphisms in an edit script. An
operational transformation of a primitive morphism $P$ w.r.t a
morphism $Q$ is another primitive morphism $P'$ such that for any
object $A:t$, $P'$ has the same \emph{effect} on $Q(A)$ as $P$ has on
$A$. We will now consider examples of mergeable types that reify the
notions of edit scripts and operational transformations.

### 2017-02-26 paper/model.tex

The \drawsome example demonstrates the utility of mergeable data types
in building concurrent applications with conceptual sharing of
state. This example, however, implicitly makes certain assumptions
about the model and the underlying system, such as the existence of a
single least common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges described in
Section~\ref{sec:semantics}. Unlike the canvas, the process of writing
merge functions for data types with non-trivial invariants could be
quite involved. The following section describes a principled approach
to deriving merge functions for any data type given its
invariants. Like \C{Canvas.merge}, merge functions often need to
compare multiple versions of a data type for structural equality,
which, when performed naively, incurs cost linear in the size of
versions. The representation of a versioned data type in \name is
optimized for performing comparisons that allows structural equality
to be determined in constant time. Such engineering details, which
contribute to the practicality of our programming model, are discussed
in Sec.~\ref{sec:impl}.

>>>>

The \drawsome example demonstrates the utility of mergeable data types
and \name programming model in building concurrent applications with
conceptual sharing of state. The model lets concurrent computations be
composed around any mergeable type. As exemplified by \drawsome,
writing a three-way \C{merge} function is the only creative process in
lifting an OCaml data type to a mergeable type. We now present few
more examples of mergeable types, and isolate a recurring pattern in
their merge functions that serves as a guide to write merge functions
for more sophisticated types.

\begin{figure}
  \centering
  \includegraphics[scale=0.4]{Figures/list-eg}

  \caption{Lists of mergeable values are mergeable. }
  \label{fig:list-eg}
\end{figure}

\input{list-code}

{\bf Lists}. List, as an abstract data type, supports three
operations: \C{insert x i l}, that inserts an element \C{x} at
position \C{i} in the list \C{l}, \C{delete i l}, that deletes the
element at \C{i}'th position, and \C{subst i x l}, that substitutes
the element at \C{i}'th position with \C{x}. Lists of mergeable items
are mergeable. The signature of a mergeable list library is shown in
Fig.~\ref{fig:mlist-sig}. Note that the type of items in the list
(\C{A.t}) is also required to be mergeable. The signature
\C{MERGEABLE} will be discussed shortly.

The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}.
\begin{itemize}
  \item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and
  computes the shortest sequence of list operations that need to be
  applied on \C{v} to obtain \C{v'}. Such a sequence is called an
  \emph{edit sequence}. The length of the sequence corresponds to the
  standard notion of the \emph{edit distance} between the two lists,
  which can be computed in polynomial time, for e.g., using the
  Wagner-Fisher algorithm~\cite{wagner-fischer}. A slight modification
  of Wagner-Fischer also lets us compute the edit sequence. The
  implementation in Fig.~\ref{fig:mlist} represents edits using the
  type \C{edit}. Constructors \C{I}, \C{D}, and \C{S} stand for
  \C{insert}, \C{delete}, and \C{subst}, respectively; \C{Nop} is
  there for technical reasons. The subst constructor also carries the
  \C{A.t} element that was substituted. Our implementation of
  Wagner-Fischer is more-or-less standard (hence, not shown), except
  that the edit sequence it returns orders \C{I}'s and \C{D}'s before
  \C{S}'s.  Fig.~\ref{fig:list-eg} illustrates edit sequences for a
  sample list. The sequence \C{[I(c,0); S(3,c,s)]} maps the list
  \C{[a;b;c]} to \C{[c;a;b;s]} because \C{subst 3 s (insert c 0
  [a;b;c]) = [c;a;b;s]}.
  \item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and
  $s_2$, that map a list $v$ to two different lists, $v_1$ and $v_2$
  (e.g., Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such
  that $s_1'$ has the same effect on $v_2$ as $s_1$ had on $v$.  For
  instance, in Fig.~\ref{list-eg}, let $s_1$ be the edit sequence
  \C{[D(1); S(d,1)]}, which maps the list \C{[a;b;c]} ($v$) to
  \C{[a;d]} ($v_1$). The \C{D} edit deletes the first element of $v$.
  However, if $s_1$ is to be applied on the list \C{[c;a;b;s]}
  ($v_2$), which has a new element at position 0, the \C{D} edit must
  delete the the second element from $v_2$, if it were to have the
  same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
  transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The
  function \C{op\_transform} of Fig.~\ref{fig:mlist-xform} computes
  such \emph{operational transformation} for the sequence \C{mine}
  w.r.t the sequence \C{others}, and returns \C{mine'} as the
  transformed sequence.  
\end{itemize}

This example, however, implicitly makes certain assumptions
about the model and the underlying system, such as the existence of a
single least common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges described in
Section~\ref{sec:semantics}. Unlike the canvas, the process of writing
merge functions for data types with non-trivial invariants could be
quite involved. The following section describes a principled approach
to deriving merge functions for any data type given its
invariants. Like \C{Canvas.merge}, merge functions often need to
compare multiple versions of a data type for structural equality,
which, when performed naively, incurs cost linear in the size of
versions. The representation of a versioned data type in \name is
optimized for performing comparisons that allows structural equality
to be determined in constant time. Such engineering details, which
contribute to the practicality of our programming model, are discussed
in Sec.~\ref{sec:impl}.
